Re-run, Repeat, Reproduce, Re-use, Replicate:\\Transforming Code into Scientific Contributions

### 2017-05-29 main.tex
conclusion

In our case, the R$^0$ version of our tiny walker seems to imply
that any version of Python would be fine.
This not the case because it uses the print {\em instruction} and the {\tt xrange} operator, both specific to Python 2.
The print {\em instruction}, available in Python 2 (still widely used, will stop to be supported in 2020), 
has been deprecated in Python 3 (first released in 2008, almost a decade ago)
in favor or a  print {\em function}.
The {\tt xrange} operator has been replaced by the {\tt range} operator in Python 3.
In order to try to future-proof the code a bit, we might as well target Python 3, as is done in the R$^1$ version.
Incidentally, it remains compatible with Python 2.

@@

\clearpage
\section*{Repeatable (R$^{\mathbf 2}$)}

@@

The code is running and producing the expected results. 
The next step is to make sure that you can produce the same output over successive runs of your program, 
should you want to. 
In other words, the next step is to make your program deterministic, 
producing {\em repeatable} output.

Such repeatability is useful. 
If a run of the program produces a particularly puzzling results, 
it allows you to scrutinize any step of the execution of the program by re-running it again. 
It is also the easiest way to prove that the program did indeed produce the published results.

Such repeatability is not always possible or easy \cite{Diethelm:2012}. But for sequential programs not depending on exotic hardware, it often comes down to controlling the initialization of the pseudo-random number generators (RNG).

@@

\clearpage
\section*{Reproducible (R$^{\mathbf 3}$)}

The R$^2$ code seems fine enough, 
but it hides several problems that come to light when trying to {\em reproduce} results.
In the strongest sense, a code is said to be {\em reproducible}
if the same results it produced in the past can be re-obtained by someone else.
As explained by \citeauthor{Mesnard:2016} \citep{Mesnard:2016}, reproducibility is harder than you think.

For instance, our R$^2$ program will not produce the same results for everyone. It will, because it is repeatable, produce the same results over repeated executions. But it will not do so over different execution environments. The cause is found in a change that occurred in the pseudo-random number generator between CPython 3.2 and CPython 3.3.

Executed with CPython 3.2 or older, the code will produce the sequence 1, 2, 1, 2, 3, 4, 5, 6, 7, 6. But with CPython 3.3 or newer, it will produce 1, 2, 1, 0, 1, 0, 1, 0, -1, 0. Reproducibility impose to specify dependencies more precisely than re-runnability: the exact versions that were used must be documented. The platform should also be described as well: CPU architecture, operating system version, endianness, etc., as well as any defaults you may have set that might have an impact. Because of the increasing complexity of our computational stacks, retrieving and deciding what is pertinent might be difficult. A best effort must be made.

Here, in addition to specifying the exact Python version and platform details, a short unit test is provided, that allows the code to verify its own reproducibility. Should this test fail, then there is little hope of reproducing the results. If it does not fail, problems may still be present.

A second problem present in the R$^2$ program is that it does not store the parameters used along with the results. Therefore, should the code change after the production of the results, someone provided with the result file will not be able to know which seed was used to produce it. In R$^3$, the seed is stored in the name of the result file.

@@

Finally, a crucial condition to making a code reproducible is making it available. As shown in \citep{Collberg:2016}, code is often not available, or available upon request. The latter may seem sufficient, but changes in email address,  career or retirement can make the code just as unavailable. Code \emph{and input and result files} should be available with the published article, as supplementary data, or through a DOI link to a scientific repository such as Figshare or Zenodo.

Providing results files is important: it allows other researchers to easily compare the results they obtains with the one used to generate the figure in the article. Comparing figures directly can be cumbersome and often only allows to conclude about a qualitative similarity.

@@

\clearpage
\section*{Reusable (R$^{\mathbf 4}$)}

@@

\clearpage
\section*{Replicable (R$^{\mathbf 5}$)}

@@

<

\clearpage
\section*{Conclusion}

>>>>

In our case, the R$^0$ version of our tiny walker seems to imply
that any version of Python would be fine.
This not the case because it uses the print {\em instruction} and the {\tt xrange} operator, both specific to Python 2.
The print {\em instruction}, available in Python 2 (still widely used, will stop to be supported in 2020), 
has been deprecated in Python 3 (first released in 2008, almost a decade ago)
in favor or a  print {\em function}.
The {\tt xrange} operator has been replaced by the {\tt range} operator in Python 3.
In order to try to future-proof the code a bit, we might as well target Python 3, as is done in the R$^1$ version.
Incidentally, it remains compatible with Python 2.\\

@@

% \clearpage
\section*{Repeatable (R$^{\mathbf 2}$)}

@@

The code is running and producing the expected results. 
The next step is to make sure that you can produce the same output over successive runs of your program, should you want to. 
In other words, the next step is to make your program deterministic,  producing {\em repeatable} output. 
Such repeatability is useful. 
If a run of the program produces a particularly puzzling results,  it allows you to scrutinize any step of the execution of the program by re-running it again. 
It is also the easiest way to prove that the program did indeed produce the published results.
Such repeatability is not always possible or easy \cite{Diethelm:2012}.
But for sequential programs not depending on exotic hardware, it often comes down to controlling the initialization of the pseudo-random number generators (RNG).\\

@@

% \clearpage
\section*{Reproducible (R$^{\mathbf 3}$)}

The R$^2$ code seems fine enough, but it hides several problems that come to light when trying to {\em reproduce} results.
In the strongest sense, a code is said to be {\em reproducible} if the same results it produced in the past can be re-obtained by someone else.
As explained by \citeauthor{Mesnard:2016} \citep{Mesnard:2016}, reproducibility is harder than you think. 
For instance, our R$^2$ program will not produce the same results for everyone. It will, because it is repeatable, produce the same results over repeated executions.
But it will not do so over different execution environments. The cause is found in a change that occurred in the pseudo-random number generator between CPython 3.2 and CPython 3.3. 
Executed with CPython 3.2 or older, the code will produce the sequence 1, 2, 1, 2, 3, 4, 5, 6, 7, 6. But with CPython 3.3 or newer, it will produce 1, 2, 1, 0, 1, 0, 1, 0, -1, 0.
Reproducibility impose to specify dependencies more precisely than re-runnability: the exact versions that were used must be documented. The platform should also be described as well: CPU architecture, operating system version, endianness, etc., as well as any defaults you may have set that might have an impact.
Because of the increasing complexity of our computational stacks, retrieving and deciding what is pertinent might be difficult. A best effort must be made.\\

Here, in addition to specifying the exact Python version and platform details, a short unit test is provided, that allows the code to verify its own reproducibility. Should this test fail, then there is little hope of reproducing the results. If it does not fail, problems may still be present. 
A second problem present in the R$^2$ program is that it does not store the parameters used along with the results. Therefore, should the code change after the production of the results, someone provided with the result file will not be able to know which seed was used to produce it. In R$^3$, the seed is stored in the name of the result file.

@@

Finally, a crucial condition to making a code reproducible is making it available. As shown in \citep{Collberg:2016}, code is often not available, or available upon request. The latter may seem sufficient, but changes in email address,  career or retirement can make the code just as unavailable. Code \emph{and input and result files} should be available with the published article, as supplementary data, or through a DOI link to a scientific repository such as Figshare or Zenodo.
Providing results files is important: it allows other researchers to easily compare the results they obtains with the one used to generate the figure in the article. Comparing figures directly can be cumbersome and often only allows to conclude about a qualitative similarity.\\

@@

% \clearpage
\section*{Re-usable (R$^{\mathbf 4}$)}

@@

% \clearpage
\section*{Replicable (R$^{\mathbf 5}$)}

@@

% \clearpage
\section*{Conclusion}

Using a very simple random walk example using Python, we've seen a variety of pitfalls of writing scientific code. The R$^0$ form, while easy to write and simple to understand is not a good scientific code while the full-fledge R$^{4}$ form is much preferred but requires a fair amount of work that might be unacceptable in some circumstances. $R^3$ form seems to be an acceptable compromise and should be accepted as the minimum scientific standard. This means this should be actually checked by reviewers and publishers when code is part of a work worth to be published. But it's hardly the case today.\\

The choice of the Python programming language to illustrate our point was not arbitrary. In less than a decade, Python has become one the few programming language of Science (together with R and Matlab). It is now used by a huge and growing number of researchers from different domains, having various experience and background in computer science. The readability and ease of use of Python make it a premium choice for new comers. But this is precisely this apparent simplicity that might hinder correct implementation. Inexperienced people may think any working code is a scientific code, but, as we've explained, this is simply not true.

% Writing reproducible software requires both knowledge and experience. For the former, there are plenty of resources online or onsite.
